---
title: "diffuStats: compute diffusion scores over networks"
author: 
-   name: Sergio Picart-Armada
    affiliation: B2SLab at Polytechnic University of Catalonia
    email: sergi.picart@upc.edu
-   name: Alexandre Perera-Lluna
    affiliation: B2SLab at Polytechnic University of Catalonia
date: "2021-02-02"
package: "`r BiocStyle::pkg_ver('diffuStats')`"
format: html
execute: 
  warning: false
  message: false
  echo: true
  eval: true
cache:
  true
---

# diffuStats: compute diffusion scores over networks

## Introduction

The general purpose `diffuStats` R package offers a collection of seven network propagation scores and five graph kernels. Those find application in ubiquitous computational biology applications, being one representative example the propagation of genetic information (e.g. disease-associated genes) in a gene-gene or a protein-protein interaction network. A distinctive feature of `diffuStats` is the implementation of statistically normalised scores, which address the recurrent question of how would the propagation of a randomised input look. It offers parametric, exact z-scores as well as permutation-based empirical probabilities.

The `diffuStats` software was published in:

> Picart-Armada, S., Thompson, W. K., Buil, A., & Perera-Lluna, A. (2018). diffuStats: an R package to compute diffusion-based scores on biological networks. Bioinformatics, 34(3), 533-534.

General guidelines on how to choose the scores, along with mathematical properties of the normalised and unnormalised scores, were published in:

> Picart-Armada, S., Thompson, W. K., Buil, A., & Perera-Lluna, A. (2020). The effect of statistical normalisation on network propagation scores. Bioinformatics, btaa896.

From versions 1.10.2/1.11.2 onwards, `diffuStats` provides functions to export the exact statistical moments (means and variances), see `?moments`. Now the users can characterise the systematic biases in the diffusion scores in their domain of application.

## Installation

`diffuStats` is part of Bioconductor, and can be installed using

```{r}
BiocManager::install("diffuStats")
```

For the development version, you can also install the package through `R CMD INSTALL` or through `devtools::install_github("b2slab/diffuStats")`, which points to its [GitHub repository](https://github.com/b2slab/diffuStats).

# Getting started

`diffuStats` is an R package providing several scores for diffusion in networks. While its original purpose lies on biological networks, its usage is not limited to that scope. In general terms, `diffuStats` builds several propagation algorithms on the \code{igraph} package [@igraph] classes and methods. A more detailed analysis and documentation of the implemented methods can be found in the protein function prediction vignette.

To get started, we will load a toy graph included in the package.

```{r}
library(diffuStats)
data("graph_toy")
```

Let's take a look in the graph:

```{r}
graph_toy
plot(graph_toy)
```

In the next section, we will be running diffusion algorithms on this tiny lattice graph.

# Specifying the input

The package `diffuStats` is flexible and allows several inputs at once for a given network. The input format is, in its most general form, a list of matrices, where each matrix contains measured nodes in rows and specific scores in columns. **Differents sets of scores may have different backgrounds**, meaning that we can specifically tag sets of nodes as **unlabelled**. If we dispose of a unique list of nodes for label propagation, we should provide a list with a unique column vector that contains `1`'s in the labels in the list and `0`'s otherwise.

In this example data, the graph contains one input already.

```{r}
input_vec <- graph_toy$input_vec

head(input_vec, 15)
```

Let's check how many nodes have values

```{r}
length(input_vec)
```

We see that all the nodes have a measure in each of the four score sets. In practice, these score sets could be disease genes, pathways, et cetera.

# The diffusion algorithm

Each one of these columns in the input can be *smoothed* using the network and new value will be derived - unlabelled nodes are also scored. This is the main purpose of diffusion: to derive new scores that intend to keep the same trends as the scores in the input, but taking into account the network structure. Equivalently, this can be regarded as a label propagation where positive and negative examples propagate their labels to their neighbouring nodes.

Let's start with the simplest case of diffusion: only a vector of values is to be smoothed. Note that these **values must be named and must be a subset or all of the graph nodes**.

```{r}
output_vec <- diffuStats::diffuse(
    graph = graph_toy, 
    method = "raw", 
    scores = input_vec)

head(output_vec, 15)
```

# Diffusion scores visualisation

The best way to visualise the scores is overlaying them in the original lattice. `diffuStats` also comes with basic mapping functions for graphical purposes. Let's see an example:

```{r}
igraph::plot.igraph(
    graph_toy, 
    vertex.color = diffuStats::scores2colours(output_vec),
    vertex.shape = diffuStats::scores2shapes(input_vec),
    main = "Diffusion scores in our lattice"
)
```

Here, we have mapped the scores to colours using `scores2colours` and we have highlighted the nodes that were in the original input using `scores2shapes` on the original scores. Square nodes were labelled as relevant in the input, and the diffusion algorithm smoothed these labels over the network - as in the guilt-by-association principle.

# Several inputs, several smoothing scores

The input to `diffuse` can be more than a vector with scores. It can be provided with a set of score vectors, stored in a matrix by columns, where rownames should contain the nodes that are being scored. As different score sets might have different labelled/unlabelled nodes, `diffuse` also accepts a list of score matrices that may have a different amount of rows.

In this section, we will diffuse using a matrix of scores that contains four sets of scores, with four different names. These example names refer to what the input contains:

-   Single: a single node is labelled as positive
-   Row: a row of nodes in the lattice graph are positives
-   Small_sample: a randomly generated small sample of the lattice nodes are positives
-   Large_sample: a randomly generated sample with half of the lattice nodes are positives

```{r}
input_mat <- graph_toy$input_mat

head(input_mat)
```

On the other hand, there are a variety of methods to compute the diffusion scores. At the moment, the following: `raw`, `ml` and `gm` for classical propagation; `z` and `mc` for scores normalised through a statistical model, and similarly `ber_s` and `ber_p`, as described in [@mosca]. The scoring methods `mc` and `ber_p` require permutations -thus being computationally intense- whereas the rest are deterministic.

For instance, let's smooth through `mc` the input matrix:

```{r}
output_mc <- diffuStats::diffuse(
    graph = graph_toy, 
    method = "mc", 
    scores = input_mat)

head(output_mc)
```

We can plot the result of the fourth column *Large_sample*:

```{r}
score_col <- 4
igraph::plot.igraph(
    graph_toy, 
    vertex.color = diffuStats::scores2colours(output_mc[, score_col]),
    vertex.shape = diffuStats::scores2shapes(input_mat[, score_col]),
    main = "Diffusion scores in our lattice"
)
```

Each method has its particularities and, in the end, it is all about the question being asked to the data and the particularities of the dataset.

# Benchmarking

Package `diffuStats` offers the option to assess the performance of the diffusion scores given user-defined target scores or labels.

The validation must be supplied with the same format as the input scores, but the labels of the nodes might be different. For example, we can diffuse labels on all the nodes of a graph but evaluate using only a specific subset of nodes and target labels. A small example: we want to evaluate how good the diffusion scores `raw` and `ml` are at recovering the original labels of the first 15 nodes when diffusing in the example network.

```{r}
df_perf <- perf(
    graph = graph_toy,
    scores = graph_toy$input_mat,
    validation = graph_toy$input_mat[1:15, ],
    grid_param = expand.grid(method = c("raw", "ml")))
df_perf
```

This indicates that both methods have a very high area under the curve in this example: the ordering of the diffusion scores is very aligned to the class label.

The last example is useful for showing a case in which diffusion scores perform poorly. As the *Small_sample* and *Large_sample* positive labels have been randomly assigned ignoring the network, diffusion is not expected to accurately predict one part of the network using as input another disjoint subset of labelled nodes. Thus, if we try to propagate the labels from nodes $1$ to $20$ and evaluate the performance using nodes from $21$ to $48$, we get a poor result:

```{r}
df_perf <- perf(
    graph = graph_toy,
    scores = graph_toy$input_mat[1:20, 3:4],
    validation = graph_toy$input_mat[21:48, 3:4],
    grid_param = expand.grid(method = c("raw", "ml")))
df_perf
```

# R session info {.unnumbered}

```{r}
sessionInfo()
```

# References

## News

File `NEWS.md` keeps track of the additions and bug fixes of each package version.
