---
title: "FELLA: Interpretation and Enrichment for Metabolomics Data"
author: "Sergio Picart-Armada [aut, cre], Francesc Fernandez-Albert [aut], Alexandre Perera-Lluna [aut]"
date: "2021-02-02"
format: html
execute: 
  warning: false
  message: false
  echo: true
  eval: false
cache:
  true
---

# Overview

FELLA is an R package designed for the enrichment of metabolomics data using KEGG entries. Given a set of affected compounds, FELLA suggests related reactions, enzymes, modules, and pathways through label propagation in a knowledge model network. The resulting subnetwork can be visualized and exported.

- This package is maintained by *Sergio Picart-Armada <sergi.picart at upc.edu>*


# Installation

To install FELLA, start R (version 4.4 or later) and enter:

```{r}

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("FELLA")
```

For older versions of R, please refer to the appropriate Bioconductor release.

# Documentation

To view documentation for the version of this package installed in your system, start R and enter:

```{r}
browseVignettes("FELLA")
```

Available vignettes include:

-   Example: a fatty liver study on *Mus musculus*
-   Example: oxybenzone exposition in gilt-head bream
-   FELLA
-   Quick start

The reference manual and news updates are also available.

# Citation:

-  From within R, enter citation("FELLA"):
-   Picart-Armada S, Fernandez-Albert F, Vinaixa M, Yanes O, Perera-Lluna A (2018). “FELLA: an R package to enrich metabolomics data.”  *BMC Bioinformatics*, **19**(1), 538. [doi:10.1186/s12859-018-2487-5](https://doi.org/10.1186/s12859-018-2487-5).
-   Picart-Armada S, Fernandez-Albert F, Vinaixa M, Rodriguez MA, Aivio S, Stracker TH, Yanes O, Perera-Lluna A (2017). “Null diffusion-based enrichment for metabolomics data.” *PLOS ONE*, **12**(12), e0189012. [doi:10.1371/journal.pone.0189012](https://doi.org/10.1371/journal.pone.0189012).

For more information, visit the [Bioconductor page for FELLA](https://bioconductor.org/packages/release/bioc/html/FELLA.html).
